---
title: "Archaeological_linear_mixed_models"
author: "DC"
date: "2024-11-30"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
  word_document:
    toc: true
    toc_depth: '4'
---
# 0. Prepare the work envrionment 
## Install necessary packages
```{r, message = FALSE, warning = FALSE}
# Install necessary packages
required_packages <- c(
  "tidyverse", "lme4", "MASS", "car", "lmerTest", "performance", 
  "splines", "brms", "elevatr", "raster", "ggplot2", "akima", 
  "fields", "sf", "sp", "RColorBrewer", "gstat", "broom.mixed"
)

# Install missing packages
install_missing <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(install_missing) > 0) {
  install.packages(install_missing)
}

# Load libraries
lapply(required_packages, library, character.only = TRUE)
```

# 1. Import the dataset and prepare the data 
```{r, message = FALSE, warning = FALSE}
ancient_chi_detail <-  read_csv("data/ancient_chinese_detail_for_mixed_effects_july.csv")
ancient_chi_detail <- ancient_chi_detail %>% 
  mutate(time_period = as.character(time_period),
         Subsistence = as.character(Subsistence))
```

## Assign the climatic data
### read the climatic data
```{r, include=FALSE}
# Load the TIFF file
Min_temp <- raster("data/wc2.1_30s_bio_6.tif")
Max_temp <- raster("data/wc2.1_30s_bio_5.tif")
Min_precip <- raster("data/wc2.1_30s_bio_14.tif")
Max_precip <- raster("data/wc2.1_30s_bio_13.tif")
```

### assign the clmatic data to groups
```{r}
coordinates_ancient_detail <- ancient_chi_detail %>%
  dplyr::select(longitude, latitude) 

# Convert to a spatial points object
points_ancient_detail <- SpatialPoints(coordinates_ancient_detail)
```

### Extract raster values for each point
```{r}
min_temp_values_ancient_detail <- extract(Min_temp, points_ancient_detail)
max_temp_values_ancient_detail <- extract(Max_temp, points_ancient_detail)
min_precip_values_ancient_detail <- extract(Min_precip, points_ancient_detail)
max_precip_values_ancient_detail <- extract(Max_precip, points_ancient_detail)
```

### Add the extracted data back to the dataset
```{r}
ancient_chi_detail  <- ancient_chi_detail  %>%
  mutate(Min_temp = min_temp_values_ancient_detail,
         Max_temp = max_temp_values_ancient_detail,
         Min_precip = min_precip_values_ancient_detail,
         Max_precip = max_precip_values_ancient_detail)
```

## Seperate males and females
```{r}
ancient_chi_male <- ancient_chi_detail %>%
  filter(sex == '0')
ancient_chi_female <- ancient_chi_detail %>%
  filter(sex == '1')
```

# 2. Influences from the environment and time periods
## Remove rows with missing values in the columns used in the models
### For FXL
```{r}
ancient_chi_male_FXL <- ancient_chi_male %>%
  filter(!is.na(FXL))
ancient_chi_female_FXL <- ancient_chi_female %>%
  filter(!is.na(FXL))
```

### For TXL
```{r}
ancient_chi_male_TXL <- ancient_chi_male %>%
  filter(!is.na(TXL))
ancient_chi_female_TXL <- ancient_chi_female %>%
  filter(!is.na(TXL))
```

### For HXL
```{r}
ancient_chi_male_HXL <- ancient_chi_male %>%
  filter(!is.na(ancient_chi_male$HXL))
ancient_chi_female_HXL <- ancient_chi_female %>%
  filter(!is.na(ancient_chi_female$HXL))
```

### For RXL
```{r}
ancient_chi_male_RXL <- ancient_chi_male %>%
  filter(!is.na(ancient_chi_male$RXL))
ancient_chi_female_RXL <- ancient_chi_female %>%
  filter(!is.na(ancient_chi_female$RXL))
```

### For FHD
```{r}
ancient_chi_male_FHD <- ancient_chi_male %>%
  filter(!is.na(FHD))
ancient_chi_female_FHD <- ancient_chi_female %>%
  filter(!is.na(FHD))
```

## 2.1 Basic Linear regression
### Fit the full linear model using lm
#### For FXL
```{r}
lm_ancient_male_fxl_full <- lm(data = ancient_chi_male_FXL, FXL ~ altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period ) # using time_period rather than subsistence to keep independence of each predictor
lm_ancient_female_fxl_full <- lm(data = ancient_chi_female_FXL, FXL ~ altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period )
```

```{r, include = FALSE}
summary(lm_ancient_male_fxl_full)
summary(lm_ancient_female_fxl_full)
```

#### For TXL
```{r}
lm_ancient_male_txl_full <- lm(data = ancient_chi_male_TXL, TXL ~ altitude_range + Min_temp + Max_temp + Min_precip + Max_precip +  time_period ) # using time_period rather than subsistence to keep independence of each predictor
lm_ancient_female_txl_full <- lm(data = ancient_chi_female_TXL, TXL ~ altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period )
```

```{r, include = FALSE}
summary(lm_ancient_male_txl_full)
summary(lm_ancient_female_txl_full)
```
#### For FHD
```{r}
lm_ancient_male_fhd_full <- lm(data = ancient_chi_male_FHD, FHD ~ altitude_range+ Min_temp + Max_temp + Min_precip + Max_precip + time_period ) # using time_period rather than subsistence to keep independence of each predictor
lm_ancient_female_fhd_full <- lm(data = ancient_chi_female_FHD, FHD ~ altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period )
```

```{r, include = FALSE}
summary(lm_ancient_male_fhd_full)
summary(lm_ancient_female_fhd_full)
```

## 2.2 Mixed effects analyis
### Fit the full mixed effects model using lmer
#### For FXL
```{r}
# Fit the model again with lmerTest loaded
lme_male_fxl_full1 <- lmer(FXL ~  altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_male_FXL)
lme_female_fxl_full1 <- lmer(FXL ~  altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_female_FXL)
```

```{r, include = FALSE}
# Now includes p-values for fixed effects
summary(lme_male_fxl_full1)  
summary(lme_female_fxl_full1)
```

#### Check the model
```{r}
check_model(lme_male_fxl_full1, check = c("linearity", "homogeneity", "normality"))
check_model(lme_female_fxl_full1, check = c("linearity", "homogeneity", "normality"))
```

#### For TXL
```{r}
# Fit the model again with lmerTest loaded
lme_male_txl_full1 <- lmer(TXL ~  altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_male_TXL)
lme_female_txl_full1 <- lmer(TXL ~  altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_female_TXL)
```

```{r, include = FALSE}
# Now includes p-values for fixed effects
summary(lme_male_txl_full1)  
summary(lme_female_txl_full1)
```

#### Check the model
```{r}
check_model(lme_male_txl_full1, check = c("linearity", "homogeneity", "normality"))
check_model(lme_female_txl_full1, check = c("linearity", "homogeneity", "normality"))
```

#### For HXL
```{r}
# Fit the model again with lmerTest loaded
lme_male_hxl_full1 <- lmer(HXL ~  altitude_range + Min_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_male_HXL)
lme_female_hxl_full1 <- lmer(HXL ~  altitude_range + Min_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_female_HXL)
```

```{r, include = FALSE}
# Now includes p-values for fixed effects
summary(lme_male_hxl_full1)  
summary(lme_female_hxl_full1)
```
#### Check the model
```{r}
check_model(lme_male_hxl_full1, check = c("linearity", "homogeneity", "normality"))
check_model(lme_female_hxl_full1, check = c("linearity", "homogeneity", "normality"))
```

#### For RXL
```{r}
# Fit the model again with lmerTest loaded
lme_male_rxl_full1 <- lmer(RXL ~  altitude_range + Min_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_male_RXL)
lme_female_rxl_full1 <- lmer(RXL ~  altitude_range + Min_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_female_RXL)
```

```{r, include = FALSE}
# Now includes p-values for fixed effects
summary(lme_male_rxl_full1)  
summary(lme_female_rxl_full1)
```

#### Check the model
```{r}
check_model(lme_male_rxl_full1, check = c("linearity", "homogeneity", "normality"))
check_model(lme_female_rxl_full1, check = c("linearity", "homogeneity", "normality"))
```

#### For FHD
```{r}
# Fit the model again with lmerTest loaded
lme_male_fhd_full1 <- lmer(FHD ~  altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_male_FHD)
lme_female_fhd_full1 <- lmer(FHD ~  altitude_range + Min_temp + Max_temp + Min_precip + Max_precip + time_period + (1 | site_id), data = ancient_chi_female_FHD)
```

```{r, include = FALSE}
# Now includes p-values for fixed effects
summary(lme_male_fhd_full1)  
summary(lme_female_fhd_full1)
```

#### Check the model
```{r}
check_model(lme_male_fhd_full1, check = c("linearity", "homogeneity", "normality"))
check_model(lme_female_fhd_full1, check = c("linearity", "homogeneity", "normality"))
```

### Plot Coefficients for full models
#### FXL
```{r, include = FALSE}
# Tidy the model outputs with specified names
tidy_male_full_fxl <- tidy(lme_male_fxl_full1, effects = "fixed")
tidy_female_full_fxl <- tidy(lme_female_fxl_full1, effects = "fixed")

# Add model labels
tidy_male_full_fxl$model <- "Male"
tidy_female_full_fxl$model <- "Female"

# Combine both male and female tidy outputs
tidy_combined_full_fxl <- rbind(tidy_male_full_fxl, tidy_female_full_fxl)

# Add significance flag to the combined data
tidy_combined_full_fxl$sig <- ifelse(tidy_combined_full_fxl$p.value < 0.05, "Significant", "Not Significant")

# Reorder terms for plotting if necessary
tidy_combined_full_fxl$term <- factor(tidy_combined_full_fxl$term, levels = unique(tidy_combined_full_fxl$term))
# Calculate confidence intervals
tidy_combined_full_fxl <- tidy_combined_full_fxl %>%
  mutate(ci_low = estimate - 1.96 * std.error,
         ci_high = estimate + 1.96 * std.error)

# Enhanced plot with significance highlighting
ggplot(tidy_combined_full_fxl, aes(x = term, y = estimate, fill = interaction(model, sig))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), 
                position = position_dodge(0.7), width = 0.2) +
  labs(x = "Predictor Variables", 
       y = "Coefficient Estimate (with 95% Confidence Interval)", 
       title = "Fixed Effects Coefficient Estimates for Male and Female Models (FXL)") +
  scale_fill_manual(values = c("Male.Significant" = "blue", 
                               "Female.Significant" = "red", 
                               "Male.Not Significant" = "lightblue", 
                               "Female.Not Significant" = "pink")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # Angle the x-axis text for readability
        plot.title = element_text(hjust = 0.5)) + # Center the plot title
  coord_flip()
```

```{r}
# Tidy the model outputs and add model labels
tidy_male_full_fxl <- tidy(lme_male_fxl_full1, effects = "fixed") %>%
  mutate(model = "Male")
tidy_female_full_fxl <- tidy(lme_female_fxl_full1, effects = "fixed") %>%
  mutate(model = "Female")

# Combine male and female outputs
tidy_combined_full_fxl <- bind_rows(tidy_male_full_fxl, tidy_female_full_fxl) %>%
  filter(term != "(Intercept)") %>% # Remove intercept
  mutate(
    sig = ifelse(p.value < 0.05, "Sig", "Not Sig"), # Add significance flag
    ci_low = estimate - 1.96 * std.error,          # Compute lower CI
    ci_high = estimate + 1.96 * std.error          # Compute upper CI
  )

# Reorder terms for plotting
tidy_combined_full_fxl$term <- factor(tidy_combined_full_fxl$term, levels = unique(tidy_combined_full_fxl$term))

# Enhanced plot with confidence intervals
plot_full_fxl_coefficients <- ggplot(tidy_combined_full_fxl, aes(x = estimate, y = term, fill = interaction(model, sig))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(xmin = ci_low, xmax = ci_high), 
                position = position_dodge(0.7), width = 0.2) +
  labs(
    y = "Predictor Variables", 
    x = "Coefficient Estimate (with 95% Confidence Interval)", 
    title = "Fixed Effects Coefficient Estimates for Male and Female Models (FXL)"
  ) +
  # Custom fill scale to distinguish significance
  scale_fill_manual(values = c(
    "Male.Sig" = "blue", 
    "Female.Sig" = "red", 
    "Male.Not Sig" = "lightblue", 
    "Female.Not Sig" = "pink"
  )) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1, size = 16), # Readable y-axis text
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16), # Readable x-axis text
    axis.title.y = element_text(size = 16), # Enlarged y-axis title
    axis.title.x = element_text(size = 16), # Enlarged x-axis title
    plot.title = element_text(hjust = 0.5, size = 18), # Centered and enlarged title
    legend.position = "bottom", # Legend at bottom
    legend.text = element_text(size = 16), # Enlarged legend text
    legend.title = element_text(size = 16) # Enlarged legend title
  ) +
  coord_flip()
```

#### TXL
```{r}
# Tidy the model outputs and add model labels
tidy_male_full_txl <- tidy(lme_male_txl_full1, effects = "fixed") %>%
  mutate(model = "Male")
tidy_female_full_txl <- tidy(lme_female_txl_full1, effects = "fixed") %>%
  mutate(model = "Female")

# Combine male and female outputs
tidy_combined_full_txl <- bind_rows(tidy_male_full_txl, tidy_female_full_txl) %>%
  filter(term != "(Intercept)") %>% # Remove intercept
  mutate(
    sig = ifelse(p.value < 0.05, "Sig", "Not Sig"),         # Add significance flag
    ci_low = estimate - 1.96 * std.error,                  # Compute lower CI
    ci_high = estimate + 1.96 * std.error                  # Compute upper CI
  )

# Reorder terms for plotting
tidy_combined_full_txl$term <- factor(tidy_combined_full_txl$term, levels = unique(tidy_combined_full_txl$term))

# Create the plot
plot_full_txl_coefficients <- ggplot(tidy_combined_full_txl, aes(y = term, x = estimate, fill = interaction(model, sig))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(xmin = ci_low, xmax = ci_high), 
                position = position_dodge(0.7), width = 0.2) +
  labs(
    y = "Predictor Variables", 
    x = "Coefficient Estimate (with 95% Confidence Interval)", 
    title = "Fixed Effects Coefficient Estimates for Male and Female Models (TXL)"
  ) +
  # Custom fill scale for significance and model interaction
  scale_fill_manual(values = c(
    "Male.Sig" = "blue", 
    "Female.Sig" = "red", 
    "Male.Not Sig" = "lightblue", 
    "Female.Not Sig" = "pink"
  )) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1, size = 16), # Readable y-axis text
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16), # Readable x-axis text
    axis.title.y = element_text(size = 16), # Enlarged y-axis title
    axis.title.x = element_text(size = 16), # Enlarged x-axis title
    plot.title = element_text(hjust = 0.5, size = 18), # Centered and enlarged title
    legend.position = "bottom", # Legend at the bottom
    legend.text = element_text(size = 16), # Enlarged legend text
    legend.title = element_text(size = 16) # Enlarged legend title
  ) +
  coord_flip()
```
#### FHD
```{r}
# Tidy the model outputs and add model labels
tidy_male_full_fhd <- tidy(lme_male_fhd_full1, effects = "fixed") %>%
  mutate(model = "Male")
tidy_female_full_fhd <- tidy(lme_female_fhd_full1, effects = "fixed") %>%
  mutate(model = "Female")

# Combine male and female outputs
tidy_combined_full_fhd <- bind_rows(tidy_male_full_fhd, tidy_female_full_fhd) %>%
  filter(term != "(Intercept)") %>% # Remove intercept
  mutate(
    sig = ifelse(p.value < 0.05, "Sig", "Not Sig"),         # Add significance flag
    ci_low = estimate - 1.96 * std.error,                  # Compute lower CI
    ci_high = estimate + 1.96 * std.error                  # Compute upper CI
  )

# Reorder terms for plotting
tidy_combined_full_fhd$term <- factor(tidy_combined_full_fhd$term, levels = unique(tidy_combined_full_fhd$term))

# Create the plot
plot_full_fhd_coefficients <- ggplot(tidy_combined_full_fhd, aes(y = term, x = estimate, fill = interaction(model, sig))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(xmin = ci_low, xmax = ci_high), 
                position = position_dodge(0.7), width = 0.2) +
  labs(
    y = "Predictor Variables", 
    x = "Coefficient Estimate (with 95% Confidence Interval)", 
    title = "Fixed Effects Coefficient Estimates for Male and Female Models (FHD)"
  ) +
  scale_fill_manual(values = c(
    "Male.Sig" = "blue", 
    "Female.Sig" = "red", 
    "Male.Not Sig" = "lightblue", 
    "Female.Not Sig" = "pink"
  )) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    plot.title = element_text(hjust = 0.5, size = 18),
    legend.position = "bottom",
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16)
  ) +
  coord_flip()
```
