---
title: "Contemporary_Chinese_groups"
author: "DC"
date: "2023-12-07"
output:
  word_document:
    toc: true
    toc_depth: '4'
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---
# Install packages
```{r, message = FALSE, warning = FALSE}
# List of required packages
required_packages <- c(
  "tidyverse",      # Data manipulation and visualization
  "sf",             # Handling spatial data
  "raster",         # Raster data operations
  "ggmap",          # Mapping and spatial visualization
  "sp",             # Spatial data classes and methods (supports sf)
  "akima",          # Interpolation of irregularly spaced data
  "rnaturalearth",  # Natural Earth map data
  "car",            # Companion to Applied Regression
  "MASS",           # Various statistical methods
  "sandwich",       # Robust covariance matrix estimators
  "lmtest",         # Testing linear regression models
  "scales",         # Scaling functions for ggplot2
  "gstat" ,         # Geostatistical modeling and kriging
  "terra"
)

# Identify missing packages and install them
missing_packages <- setdiff(required_packages, rownames(installed.packages()))
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load all required libraries
invisible(lapply(required_packages, library, character.only = TRUE))
```

# 1. Prepare the data
Read the group data on contemporary populations
```{r, message = FALSE}
groups <- read_csv("data/modern_multiple_groups.csv")
```

## Assign the climatic data
### Read the climatic data
```{r}
# Load the TIFF file
Min_temp <- raster("data/wc2.1_30s_bio_6.tif")
Max_temp <- raster("data/wc2.1_30s_bio_5.tif")
Min_precip <- raster("data/wc2.1_30s_bio_14.tif")
Max_precip <- raster("data/wc2.1_30s_bio_13.tif")
```

### Assign the clmatic data to modern groups
```{r}
coordinates <- groups %>%
  dplyr::select(`Longitude (E)`, `Latitude (N)`)  # Use backticks around column names with spaces

# Convert to a spatial points object
points <- SpatialPoints(coordinates)
```

```{r}
# Extract raster values for each point
min_temp_values <- extract(Min_temp, points)
max_temp_values <- extract(Max_temp, points)
min_precip_values <- extract(Min_precip, points)
max_precip_values <- extract(Max_precip, points)
```

```{r}
# Add the extracted data back to the dataset
groups <- groups %>%
  mutate(Min_temp = min_temp_values,
         Max_temp = max_temp_values,
         Min_precip = min_precip_values,
         Max_precip = max_precip_values)
```

## Get China Map
```{r, message =FALSE}
shapefile_path <- "data/bou1_4p.shp"
# Read the shapefile
china_shape <- st_read(shapefile_path)
```

# 2. Regression analysis
## Separate data by sex
```{r}
# Use dplyr::filter() to avoid conflicts
male_data <- dplyr::filter(groups, Sex == "M")
female_data <- dplyr::filter(groups, Sex == "F")

# Convert necessary columns to numeric
male_data <- male_data %>%
  mutate(Stature = as.numeric(Stature_cm),
         Body_mass = as.numeric(Body_mass_kg),
         Sitting_height = as.numeric(R_Sitting_ht),
         Min_temp = as.numeric(Min_temp),
         Max_temp = as.numeric(Max_temp),
         Min_precip = as.numeric(Min_precip),
         Max_precip = as.numeric(Max_precip))


female_data <- female_data %>%
  mutate(Stature = as.numeric(Stature_cm),
         Body_mass = as.numeric(Body_mass_kg),
         Sitting_height = as.numeric(R_Sitting_ht),
         Min_temp = as.numeric(Min_temp),
         Max_temp = as.numeric(Max_temp),
         Min_precip = as.numeric(Min_precip),
         Max_precip = as.numeric(Max_precip))
```

## 2.1 Ordinary linear regression
```{r}
# Fit linear models for males
male_stature_lm <- lm(Stature ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range + Time, data = male_data)
male_sitting_lm <- lm(Sitting_height ~ Min_temp + Max_temp + Min_precip + Max_precip +Altitude_range+ Time, data = male_data)
male_bodymass_lm <- lm(Body_mass ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range+ Time, data = male_data)

# Fit linear models for females
female_stature_lm <- lm(Stature ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range+ Time, data = female_data)
female_sitting_lm <- lm(Sitting_height ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range+ Time, data = female_data)
female_bodymass_lm <- lm(Body_mass ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range+ Time, data = female_data)
```
```{r}
# Function to check model assumptions
check_assumptions <- function(model) {
  par(mfrow=c(2,2))  # Set up plot area for 4 diagnostic plots

  # 1. Residuals vs Fitted: Check for linearity
  plot(model, which=1, main="Residuals vs Fitted")

  # 2. Normal Q-Q: Check for normality of residuals
  plot(model, which=2, main="Normal Q-Q")

  # 3. Scale-Location: Check for homoscedasticity (constant variance)
  plot(model, which=3, main="Scale-Location")

  # 4. Residuals vs Leverage: Check for influential points
  plot(model, which=5, main="Residuals vs Leverage")
}

# Function to check multicollinearity
check_vif <- function(model) {
  vif(model)  # VIF values above 5 indicate multicollinearity
}

# Check model assumptions and multicollinearity for males
check_assumptions(male_stature_lm)
check_assumptions(male_sitting_lm)
check_assumptions(male_bodymass_lm)

check_vif(male_stature_lm)
check_vif(male_sitting_lm)
check_vif(male_bodymass_lm)

# Check model assumptions and multicollinearity for females
check_assumptions(female_stature_lm)
check_assumptions(female_sitting_lm)
check_assumptions(female_bodymass_lm)

check_vif(female_stature_lm)
check_vif(female_sitting_lm)
check_vif(female_bodymass_lm)
```

## 2.2 Robust regression
```{r}
# Robust regression for male models
male_stature_rlm <- rlm(Stature ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range +Time, data = male_data)
male_sitting_rlm <- rlm(Sitting_height ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range +Time, data = male_data)
male_bodymass_rlm <- rlm(Body_mass ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range +Time, data = male_data)

# Robust regression for female models
female_stature_rlm <- rlm(Stature ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range +Time, data = female_data)
female_sitting_rlm <- rlm(Sitting_height ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range +Time, data = female_data)
female_bodymass_rlm <- rlm(Body_mass ~ Min_temp + Max_temp + Min_precip + Max_precip + Altitude_range+Time, data = female_data)
```

### Plot coefficients
```{r}
# Combine male and female data for plotting
combined_data <- rbind(
  data.frame(male_combined_data, Sex = "Male"),
  data.frame(female_combined_data, Sex = "Female")
)

# Add a column for asterisks to indicate significant predictors
combined_data$Significant_Asterisk <- ifelse(combined_data$Significance == "Significant", "*", "")

combined_plot <- ggplot(combined_data, aes(y = Coefficient, x = Variable)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", fill = "lightblue") +  # Uniform fill color
  
  geom_errorbar(aes(ymin = Coefficient - 1.96 * Std_Error, ymax = Coefficient + 1.96 * Std_Error), width = 0.2, position = position_dodge(0.9)) +
  geom_text(aes(label = Significant_Asterisk), color = "red", size = 5, vjust = -0.5, position = position_dodge(width = 0.9)) +
  facet_grid(Model ~ Sex) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  # Horizontal line at 0 on the y-axis
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "bottom") +  # Adjusted axis text for readability
  labs(title = "Male and Female Model Coefficients (Robust Regression)", y = "Coefficient", x = "Variable")

# Display the combined plot
print(combined_plot)
```
  
  # 4. Print summary for robust linear models (rlm)
  if (inherits(model, "rlm")) {
    cat("Summary for robust linear model (RLM):\n")
    print(summary(model))
  }
  
  cat("\nModel diagnostics completed.\n")
}

# Diagnose a standard linear model:
 diagnose_model(male_stature_lm)

 # Diagnose a robust linear model:
 diagnose_model(male_stature_rlm)

# Diagnose with robust standard errors for a standard linear model:
 diagnose_model(male_stature_lm, robust_se = TRUE)
```

## 2.3 Compare the Han, Sherpa and Tibetan Groups
```{r}
# Set the desired order for the Ethnic_Group factor
filtered_data$Ethnic_Group <- factor(filtered_data$Ethnic_Group, levels = c("Han", "Tibetan", "Sherpa"))
# Convert Longitude and Latitude to numeric
filtered_data$Longitude_E <- as.numeric(filtered_data$Longitude_E)
filtered_data$Latitude_N <- as.numeric(filtered_data$Latitude_N)
```

```{r}
# Set the desired order for the Ethnic_Group factor, including all groups
groups$Ethnic_Group <- factor(groups$Ethnic_Group, levels = unique(groups$Ethnic_Group))

# Create a new column named 'ethnic_group_2' for the desired classifications
groups <- groups %>%
  mutate(ethnic_group_2 = case_when(
    Ethnic_Group == "Tibetan" ~ "Tibetan",
    Ethnic_Group == "Sherpa" ~ "Sherpa",
    Ethnic_Group == "Han" & `Latitude (N)` > 34 ~ "Han (Northern)",
    Ethnic_Group == "Han" & `Longitude (E)` <= 34 ~ "Han (Southern)",
    TRUE ~ NA_character_
  ))
```

```{r, warning = FALSE}
# Plot with merged time periods for Body Mass with linear regression line using 'ethnic_group_2'
body_mass_plot <- ggplot(groups_filtered, aes(x = Stature_cm, y = Body_mass_kg, color = Sex, shape = ethnic_group_2, linetype = ethnic_group_2)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line
  labs(x = "Stature (cm)", y = "Body Mass", title = "Stature vs Body Mass by Ethnic Group and Time Period", shape = "Ethnic group", linetype = "Group") +
  theme(legend.position = "bottom") +
  facet_grid(Sex ~ Time_period_merged) +
  scale_shape_manual(values = c(1, 2, 3, 4)) +  # Set different shapes for ethnic groups
  scale_linetype_manual(values = c("solid", "dashed","twodash", "blank")) +  # Set different line types for ethnic groups
  guides(shape = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2), color = guide_legend(nrow = 2))
body_mass_plot
```

# 3 Interpolation over time and space
## 3.1 Stature
```{r}
# Convert columns to numeric, replacing non-numeric values with NA
convert_to_numeric <- function(df, columns) {
  for (col in columns) {
    df[[col]] <- as.numeric(df[[col]])
  }
  return(df)
}

# Apply the function to convert 'Body_mass(kg)' and 'R_Sitting_ht' to numeric
groups <- convert_to_numeric(groups, c("Body_mass(kg)", "R_Sitting_ht"))

# Function to subset data based on sex and time period
subset_data <- function(df, sex, time_periods) {
  subset(df, Sex == sex & Time_period %in% time_periods,
         select = c(`Latitude (N)`, `Longitude (E)`, `Stature(cm)`, `Body_mass(kg)`, R_Sitting_ht, 
                    Ethnic_Group, Time, Altitude_range, Relative_Living_Bi_iliaca_breadth))
}

# Subset data for each group
ht_china_male_early <- subset_data(groups, "M", c(1, 2))
ht_china_female_early <- subset_data(groups, "F", c(1, 2))
ht_china_male_late <- subset_data(groups, "M", c(3))
ht_china_female_late <- subset_data(groups, "F", c(3))
```

```{r, warning = FALSE}
# Function to rename columns for all datasets
rename_columns <- function(df) {
  names(df) <- c("Latitude_N", "Longitude_E", "Stature_cm", "Body_mass_kg", "R_Sitting_ht","Ethnic_Group", "Time", "Altitude_range")
  return(df)
}

# Apply the rename function to all datasets
ht_china_male_early <- rename_columns(ht_china_male_early)
ht_china_female_early <- rename_columns(ht_china_female_early)
ht_china_male_late <- rename_columns(ht_china_male_late)
ht_china_female_late <- rename_columns(ht_china_female_late)
```

```{r}
# Define the IDW interpolation function
perform_idw_interpolation <- function(df, filename) {
  # Check if the dataframe has coordinates set properly
  if (!("Longitude_E" %in% names(df)) || !("Latitude_N" %in% names(df)) || !("Stature_cm" %in% names(df))) {
    stop("The data frame must contain 'Longitude_E', 'Latitude_N', and 'Stature_cm' columns.")
  }
  
  # Define the grid extent
  xmin <- 70
  xmax <- 140
  ymin <- 6
  ymax <- 60

  # Set the number of cells in x and y dimensions to ensure equal cell size
  cell_size <- (xmax - xmin) / 400  # Assuming 400 cells in the x-direction

  # Create raster with equal horizontal and vertical resolution
  htrChina <- raster(
    xmn = xmin, xmx = xmax,
    ymn = ymin, ymx = ymax,
    res = cell_size,  # Specify resolution to be the same in both x and y
    crs = "+proj=longlat +datum=WGS84"
  )

  # Convert the raster to points to extract the coordinates
  htrChina_points <- rasterToPoints(htrChina, spatial = TRUE)

  # Set spatial coordinates for the input data frame
  coordinates(df) <- ~ Longitude_E + Latitude_N

  # Calculate distances between observation points and raster grid points
  dists <- spDists(
    x = coordinates(df),
    y = coordinates(htrChina_points),
    longlat = TRUE
  )

  # Inverse distance weighted interpolation
  idp <- 2
  inv.w <- (1 / (dists ^ idp))
  z <- (t(inv.w) %*% matrix(df$Stature_cm)) / apply(inv.w, 2, sum)

  # Update the raster with predicted values
  htpred.China <- htrChina
  values(htpred.China) <- z

  # Export the data as a raster file in ASCII format
  writeRaster(htpred.China, filename = filename, format = "ascii", overwrite = TRUE)
}

# Apply the function to each group
# Interpolation for Early Female Population
perform_idw_interpolation(ht_china_male_early, "HT_IDW2_China_male_early.asc")

perform_idw_interpolation(ht_china_female_early, "HT_IDW2_China_female_early.asc")

# Interpolation for Late Male Population
perform_idw_interpolation(ht_china_male_late, "HT_IDW2_China_male_late.asc")

# Interpolation for Late Female Population
perform_idw_interpolation(ht_china_female_late, "HT_IDW2_China_female_late.asc")
```

```{r}
htpred_male_early <- raster("HT_IDW2_China_male_early.asc")
htpred_female_early <- raster("HT_IDW2_China_female_early.asc")
htpred_male_late <- raster("HT_IDW2_China_male_late.asc")
htpred_female_late <- raster("HT_IDW2_China_female_late.asc")

# Convert the rasters to data frames
htpred_male_early_df <- as.data.frame(htpred_male_early, xy = TRUE)
colnames(htpred_male_early_df) <- c("Longitude", "Latitude", "Stature")

htpred_female_early_df <- as.data.frame(htpred_female_early, xy = TRUE)
colnames(htpred_female_early_df) <- c("Longitude", "Latitude", "Stature")

htpred_male_late_df <- as.data.frame(htpred_male_late, xy = TRUE)
colnames(htpred_male_late_df) <- c("Longitude", "Latitude", "Stature")

htpred_female_late_df <- as.data.frame(htpred_female_late, xy = TRUE)
colnames(htpred_female_late_df) <- c("Longitude", "Latitude", "Stature")
```

```{r}
# Function to plot raster heatmap overlaid on the China base map with dataset points by Ethnic Group using shapes and colors
plot_heatmap_with_points_on_china_map <- function(raster_df, points_df, title) {
  # Determine the number of unique ethnic groups
  unique_ethnic_groups <- unique(points_df$Ethnic_Group)
  num_groups <- length(unique_ethnic_groups)

  # Define shape values for each ethnic group
  # Assign specific shapes to Han and Tibetan groups
  shape_values <- rep(c(15:22), length.out = num_groups)
  names(shape_values) <- unique_ethnic_groups
  shape_values["Han"] <- 17      # Unify the Han group shape to a triangle
  shape_values["Tibetan"] <- 16  # Unify the Tibetan group shape to a circle
shape_values["Sherpa"] <- 15  # Unify the Tibetan group shape to a circle
  ggplot() +
    geom_sf(data = china_shape, fill = "white", color = "black") +  # Plot the base map of China
    geom_raster(data = raster_df, aes(x = Longitude, y = Latitude, fill = Stature), alpha = 0.7) +  # Plot the heatmap
    geom_point(data = points_df, aes(x = Longitude_E, y = Latitude_N, shape = Ethnic_Group, color = Ethnic_Group), size = 1) +  # Plot points using different shapes and colors
    scale_shape_manual(values = shape_values) +  # Assign custom shapes, including unified Han and Tibetan group shapes
    scale_color_viridis_d(option = "plasma") +  # Assign colors using viridis color palette (option 'plasma' for variety)
    scale_fill_viridis_c() +  # Use viridis color scale for the heatmap
    theme_minimal() +
    labs(title = title,
         x = "Longitude",
         y = "Latitude",
         fill = "Stature",
         shape = "Ethnic Group",
         color = "Ethnic Group") +  # Include color legend for further differentiation
    coord_sf()  # Keep the coordinate reference consistent
}

# Plotting each dataset separately with the heatmap and the China base map

# Plot heatmap for Early Male Population with points
plot1 <- plot_heatmap_with_points_on_china_map(htpred_male_early_df, ht_china_male_early, "Heatmap with Points - Early Male Population")

# Plot heatmap for Early Female Population with points
plot2 <- plot_heatmap_with_points_on_china_map(htpred_female_early_df, ht_china_female_early, "Heatmap with Points - Early Female Population")

# Plot heatmap for Late Male Population with points
plot3 <- plot_heatmap_with_points_on_china_map(htpred_male_late_df, ht_china_male_late, "Heatmap with Points - Late Male Population")

# Plot heatmap for Late Female Population with points
plot4 <- plot_heatmap_with_points_on_china_map(htpred_female_late_df, ht_china_female_late, "Heatmap with Points - Late Female Population")
```

```{r}
# Set global min and max stature values across datasets for consistent heatmap scale
global_min <- min(htpred_female_early_df$Stature, na.rm = TRUE)
global_max <- max(htpred_male_late_df$Stature, na.rm = TRUE)

# Function to plot raster heatmap overlaid on the China base map with dataset points by Ethnic Group using shapes and colors
plot_heatmap_with_points_on_china_map <- function(raster_df, points_df, title, group_name) {
  # Determine the number of unique ethnic groups
  unique_ethnic_groups <- unique(points_df$Ethnic_Group)
  num_groups <- length(unique_ethnic_groups)

  # Define shape values for each ethnic group
  # Assign specific shapes to Han, Tibetan, and Sherpa groups
  shape_values <- rep(c(1:14), length.out = num_groups)
  names(shape_values) <- unique_ethnic_groups
  shape_values["Han"] <- 15     # Unify the Han group shape 
  shape_values["Tibetan"] <-19  # Unify the Tibetan group shape
  shape_values["Sherpa"] <- 17 # Assign specific shape for Sherpa group
  shape_values["Mongols"] <- 18 # Assign specific shape for Mongols group
  shape_values["Zhuang"] <- 20 # Assign specific shape for Zhuang group
  shape_values["Uyghurs"] <- 21 # Assign specific shape for Uyghurs group
  shape_values["Gelao"] <- 22 # Assign specific shape for Gelao group
  # Create the plot with unified layout settings
  p <- ggplot() +
    geom_sf(data = china_shape, fill = "white", color = "black") +  # Plot the base map of China
    geom_raster(data = raster_df, aes(x = Longitude, y = Latitude, fill = Stature), alpha = 0.7) +  # Plot the heatmap
    geom_point(data = points_df, aes(x = Longitude_E, y = Latitude_N, shape = Ethnic_Group, color = Ethnic_Group), size = 1) +  # Plot points using different shapes and colors
    scale_shape_manual(values = shape_values) +  # Assign custom shapes, including unified Han, Tibetan, and Sherpa group shapes
    scale_color_viridis_d(option = "plasma") +  # Assign colors using viridis color palette (option 'plasma' for variety)
    scale_fill_viridis_c(limits = c(global_min, global_max)) +  # Use viridis color scale for the heatmap with unified limits
    theme_minimal() +
    labs(title = title,
         x = "Longitude",
         y = "Latitude",
         fill = "Stature(cm)",
         shape = "Ethnic Group",
         color = "Ethnic Group",
         caption = group_name) +  # Add the group name under the plot
    theme(
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  # Center the title and make it bold
      axis.title = element_text(size = 10),  # Standardize axis title font size
      axis.text = element_text(size = 10),   # Standardize axis label size
      legend.position = "right",              # Move the legend (scale) to the left
      legend.title = element_text(size = 10),# Standardize legend title size
      legend.text = element_text(size = 8), # Standardize legend text size
      plot.margin = unit(c(1, 1, 1, 1), "cm") # Set consistent plot margins
    ) +
    coord_sf()  # Keep the coordinate reference consistent

  return(p)
}

# Plotting each dataset separately with the heatmap and the China base map

# Plot heatmap for Early Male Population with points
plot1 <- plot_heatmap_with_points_on_china_map(htpred_male_early_df, ht_china_male_early, "Heatmap with Points - Early Male Population", "Early Male Population")

# Plot heatmap for Early Female Population with points
plot2 <- plot_heatmap_with_points_on_china_map(htpred_female_early_df, ht_china_female_early, "Heatmap with Points - Early Female Population", "Early Female Population")

# Plot heatmap for Late Male Population with points
plot3 <- plot_heatmap_with_points_on_china_map(htpred_male_late_df, ht_china_male_late, "Heatmap with Points - Late Male Population", "Late Male Population")

# Plot heatmap for Late Female Population with points
plot4 <- plot_heatmap_with_points_on_china_map(htpred_female_late_df, ht_china_female_late, "Heatmap with Points - Late Female Population", "Late Female Population")
```
## 3.2 Body Mass
```{r}
# Convert columns to numeric, replacing non-numeric values with NA
convert_to_numeric <- function(df, columns) {
  for (col in columns) {
    df[[col]] <- suppressWarnings(as.numeric(df[[col]]))  # Suppress warnings to avoid output clutter
  }
  return(df)
}

# Apply the function to convert columns to numeric
groups <- convert_to_numeric(groups, c("Body_mass(kg)", "R_Sitting_ht"))

# Function to subset data based on sex and time period
subset_data <- function(df, sex, time_periods) {
  subset(df, Sex == sex & Time_period %in% time_periods,
         select = c(`Latitude (N)`, `Longitude (E)`, `Stature(cm)`, `Body_mass(kg)`, `R_Sitting_ht`, 
                    Ethnic_Group, Time, Altitude_range, `Relative_Living_Bi_iliaca_breadth`))
}

# Subset data for each group
ht_china_male_early <- subset_data(groups, "M", c(1, 2))
ht_china_female_early <- subset_data(groups, "F", c(1, 2))
ht_china_male_late <- subset_data(groups, "M", c(3))
ht_china_female_late <- subset_data(groups, "F", c(3))

# Function to rename columns for all datasets
rename_columns <- function(df) {
  names(df) <- c("Latitude_N", "Longitude_E", "Stature_cm", "Body_mass_kg", "R_Sitting_ht", 
                 "Ethnic_Group", "Time", "Altitude_range", "Relative_Living_Bi_iliaca_breadth")
  return(df)
}

# Apply the rename function to all datasets
ht_china_male_early <- rename_columns(ht_china_male_early)
ht_china_female_early <- rename_columns(ht_china_female_early)
ht_china_male_late <- rename_columns(ht_china_male_late)
ht_china_female_late <- rename_columns(ht_china_female_late)

# Function to set spatial coordinates and plot points for each dataset
plot_points <- function(df, title, color_function = heat.colors) {
  # Set spatial coordinates
  coordinates(df) <- ~ Longitude_E + Latitude_N
  
  # Create a plot for the current dataset using a color function
  df$color_index <- as.numeric(cut(df$Body_mass_kg, breaks = 10))  # Categorize Body_mass_kg into 10 bins
  plot(df, col = color_function(10)[df$color_index], pch = 16, cex = 0.6, main = title)
}

# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Plot each dataset
plot_points(ht_china_male_early, title = "Early Male Population")
plot_points(ht_china_female_early, title = "Early Female Population")
plot_points(ht_china_male_late, title = "Late Male Population")
plot_points(ht_china_female_late, title = "Late Female Population")
```

```{r}
# Define the IDW interpolation function
perform_idw_interpolation <- function(df, filename) {
  # Check if the dataframe has coordinates set properly
  if (!("Longitude_E" %in% names(df)) || !("Latitude_N" %in% names(df)) || !("Body_mass_kg" %in% names(df))) {
    stop("The data frame must contain 'Longitude_E', 'Latitude_N', and 'Body_mass_kg' columns.")
  }
  
  # Remove rows with NA values in relevant columns
  df <- df[!is.na(df$Longitude_E) & !is.na(df$Latitude_N) & !is.na(df$Body_mass_kg), ]
  
  # Define the grid extent
  xmin <- 70
  xmax <- 140
  ymin <- 6
  ymax <- 60

  # Set the number of cells in x and y dimensions to ensure equal cell size
  cell_size <- (xmax - xmin) / 400  # Assuming 400 cells in the x-direction

  # Create raster with equal horizontal and vertical resolution
  htrChina <- raster(
    xmn = xmin, xmx = xmax,
    ymn = ymin, ymx = ymax,
    res = cell_size,  # Specify resolution to be the same in both x and y
    crs = "+proj=longlat +datum=WGS84"
  )

  # Convert the raster to points to extract the coordinates
  htrChina_points <- rasterToPoints(htrChina, spatial = TRUE)

  # Set spatial coordinates for the input data frame
  coordinates(df) <- ~ Longitude_E + Latitude_N

  # Calculate distances between observation points and raster grid points
  dists <- spDists(
    x = coordinates(df),
    y = coordinates(htrChina_points),
    longlat = TRUE
  )

  # Inverse distance weighted interpolation
  idp <- 2
  inv.w <- (1 / (dists ^ idp))
  z <- (t(inv.w) %*% matrix(df$Body_mass_kg)) / apply(inv.w, 2, sum)

  # Update the raster with predicted values
  htpred.China <- htrChina
  values(htpred.China) <- z

  # Export the data as a raster file in ASCII format
  writeRaster(htpred.China, filename = filename, format = "ascii", overwrite = TRUE)
}

# Apply the function to each group
perform_idw_interpolation(ht_china_male_early, "HT_IDW2_China_male_early.asc")

# Interpolation for Early Female Population
perform_idw_interpolation(ht_china_female_early, "HT_IDW2_China_female_early.asc")

# Interpolation for Late Male Population
perform_idw_interpolation(ht_china_male_late, "HT_IDW2_China_male_late.asc")

# Interpolation for Late Female Population
perform_idw_interpolation(ht_china_female_late, "HT_IDW2_China_female_late.asc")
```

```{r}
# Load the interpolated rasters
htpred_male_early <- raster("HT_IDW2_China_female_early.asc")
htpred_female_early <- raster("HT_IDW2_China_female_early.asc")
htpred_male_late <- raster("HT_IDW2_China_male_late.asc")
htpred_female_late <- raster("HT_IDW2_China_female_late.asc")

# Convert the rasters to data frames
htpred_male_early_df <- as.data.frame(htpred_male_early, xy = TRUE)
colnames(htpred_male_early_df) <- c("Longitude", "Latitude", "Body_mass_kg")

htpred_female_early_df <- as.data.frame(htpred_female_early, xy = TRUE)
colnames(htpred_female_early_df) <- c("Longitude", "Latitude", "Body_mass_kg")

htpred_male_late_df <- as.data.frame(htpred_male_late, xy = TRUE)
colnames(htpred_male_late_df) <- c("Longitude", "Latitude", "Body_mass_kg")

htpred_female_late_df <- as.data.frame(htpred_female_late, xy = TRUE)
colnames(htpred_female_late_df) <- c("Longitude", "Latitude", "Body_mass_kg")
```

```{r}
# Calculate global min and max for Body_mass_kg across all datasets to unify the scale
global_min <- min(c(min(htpred_male_early_df $Body_mass_kg, na.rm = TRUE),
                    min(htpred_female_early_df$Body_mass_kg, na.rm = TRUE),
                    min(htpred_male_late_df$Body_mass_kg, na.rm = TRUE),
                    min(htpred_female_late_df$Body_mass_kg, na.rm = TRUE)))

global_max <- max(c(max(htpred_male_early_df$Body_mass_kg, na.rm = TRUE),
                    max(htpred_female_early_df$Body_mass_kg, na.rm = TRUE),
                    max(htpred_male_late_df$Body_mass_kg, na.rm = TRUE),
                    max(htpred_female_late_df$Body_mass_kg, na.rm = TRUE)))
```

```{r}
# Function to plot raster heatmap overlaid on the China base map with dataset points by Ethnic Group using shapes and colors
plot_heatmap_with_points_on_china_map <- function(raster_df, points_df, title, group_name) {
  # Determine the number of unique ethnic groups
  unique_ethnic_groups <- unique(points_df$Ethnic_Group)
  num_groups <- length(unique_ethnic_groups)

  # Define shape values for each ethnic group
  # Assign specific shapes to Han, Tibetan, and Sherpa groups
  shape_values <- rep(c(1:14), length.out = num_groups)
  names(shape_values) <- unique_ethnic_groups
  shape_values["Han"] <- 15     # Unify the Han group shape 
  shape_values["Tibetan"] <-19  # Unify the Tibetan group shape
  shape_values["Sherpa"] <- 17 # Assign specific shape for Sherpa group
  shape_values["Mongols"] <- 18 # Assign specific shape for Mongols group
  shape_values["Zhuang"] <- 20 # Assign specific shape for Zhuang group
  shape_values["Uyghurs"] <- 21 # Assign specific shape for Uyghurs group
  shape_values["Gelao"] <- 22 # Assign specific shape for Gelao group
  # Create the plot with unified layout settings
  p <- ggplot() +
    geom_sf(data = china_shape, fill = "white", color = "black") +  # Plot the base map of China
    geom_raster(data = raster_df, aes(x = Longitude, y = Latitude, fill = Body_mass_kg), alpha = 0.7) +  # Plot the heatmap
    geom_point(data = points_df, aes(x = Longitude_E, y = Latitude_N, shape = Ethnic_Group, color = Ethnic_Group), size = 1) +  # Plot points using different shapes and colors
    scale_shape_manual(values = shape_values) +  # Assign custom shapes, including unified Han, Tibetan, and Sherpa group shapes
    scale_color_viridis_d(option = "plasma") +  # Assign colors using viridis color palette (option 'plasma' for variety)
    scale_fill_viridis_c(limits = c(global_min, global_max)) +  # Use viridis color scale for the heatmap with unified limits
    theme_minimal() +
    labs(title = title,
         x = "Longitude",
         y = "Latitude",
         fill = "Body mass (kg)",
         shape = "Ethnic Group",
         color = "Ethnic Group",
         caption = group_name) +  # Add the group name under the plot
    theme(
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  # Center the title and make it bold
      axis.title = element_text(size = 10),  # Standardize axis title font size
      axis.text = element_text(size = 10),   # Standardize axis label size
      legend.position = "right",              # Move the legend (scale) to the left
      legend.title = element_text(size = 10),# Standardize legend title size
      legend.text = element_text(size = 8), # Standardize legend text size
      plot.margin = unit(c(1, 1, 1, 1), "cm") # Set consistent plot margins
    ) +
    coord_sf()  # Keep the coordinate reference consistent

  return(p)
}

# Plotting each dataset separately with the heatmap and the China base map

# Plot heatmap for Early Male Population with points
plot1 <- plot_heatmap_with_points_on_china_map(htpred_male_early_df, ht_china_male_early, "Heatmap with Points - Early Male Population", "Early Male Population")

# Plot heatmap for Early Female Population with points
plot2 <- plot_heatmap_with_points_on_china_map(htpred_female_early_df, ht_china_female_early, "Heatmap with Points - Early Female Population", "Early Female Population")

# Plot heatmap for Late Male Population with points
plot3 <- plot_heatmap_with_points_on_china_map(htpred_male_late_df, ht_china_male_late, "Heatmap with Points - Late Male Population", "Late Male Population")

# Plot heatmap for Late Female Population with points
plot4 <- plot_heatmap_with_points_on_china_map(htpred_female_late_df, ht_china_female_late, "Heatmap with Points - Late Female Population", "Late Female Population")
```

## 3.3 Sitting height
```{r}
# Convert columns to numeric, replacing non-numeric values with NA
convert_to_numeric <- function(df, columns) {
  for (col in columns) {
    df[[col]] <- suppressWarnings(as.numeric(df[[col]]))  # Suppress warnings to avoid output clutter
  }
  return(df)
}

# Apply the function to convert columns to numeric
groups <- convert_to_numeric(groups, c("Body_mass(kg)", "R_Sitting_ht"))

# Function to subset data based on sex and time period
subset_data <- function(df, sex, time_periods) {
  subset(df, Sex == sex & Time_period %in% time_periods,
         select = c(`Latitude (N)`, `Longitude (E)`, `Stature(cm)`, `Body_mass(kg)`, `R_Sitting_ht`, 
                    Ethnic_Group, Time, Altitude_range, `Relative_Living_Bi_iliaca_breadth`))
}

# Subset data for each group
ht_china_male_early <- subset_data(groups, "M", c(1, 2))
ht_china_female_early <- subset_data(groups, "F", c(1, 2))
ht_china_male_late <- subset_data(groups, "M", c(3))
ht_china_female_late <- subset_data(groups, "F", c(3))

# Function to rename columns for all datasets
rename_columns <- function(df) {
  names(df) <- c("Latitude_N", "Longitude_E", "Stature_cm", "Body_mass_kg", "R_Sitting_ht", 
                 "Ethnic_Group", "Time", "Altitude_range", "Relative_Living_Bi_iliaca_breadth")
  return(df)
}

# Apply the rename function to all datasets
ht_china_male_early <- rename_columns(ht_china_male_early)
ht_china_female_early <- rename_columns(ht_china_female_early)
ht_china_male_late <- rename_columns(ht_china_male_late)
ht_china_female_late <- rename_columns(ht_china_female_late)

# Function to set spatial coordinates and plot points for each dataset
plot_points <- function(df, title, color_function = heat.colors) {
  # Set spatial coordinates
  coordinates(df) <- ~ Longitude_E + Latitude_N
  
  # Create a plot for the current dataset using a color function
  df$color_index <- as.numeric(cut(df$R_Sitting_ht, breaks = 10))  # Categorize R_Sitting_ht into 10 bins
  plot(df, col = color_function(10)[df$color_index], pch = 16, cex = 0.6, main = title)
}

# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Plot each dataset
plot_points(ht_china_male_early, title = "Early Male Population")
plot_points(ht_china_female_early, title = "Early Female Population")
plot_points(ht_china_male_late, title = "Late Male Population")
plot_points(ht_china_female_late, title = "Late Female Population")
```

```{r}
# Define the IDW interpolation function
perform_idw_interpolation <- function(df, filename) {
  # Check if the dataframe has coordinates set properly
  if (!("Longitude_E" %in% names(df)) || !("Latitude_N" %in% names(df)) || !("R_Sitting_ht" %in% names(df))) {
    stop("The data frame must contain 'Longitude_E', 'Latitude_N', and 'R_Sitting_ht' columns.")
  }
  
  # Remove rows with NA values in relevant columns
  df <- df[!is.na(df$Longitude_E) & !is.na(df$Latitude_N) & !is.na(df$R_Sitting_ht), ]
  
  # Define the grid extent
  xmin <- 70
  xmax <- 140
  ymin <- 6
  ymax <- 60

  # Set the number of cells in x and y dimensions to ensure equal cell size
  cell_size <- (xmax - xmin) / 400  # Assuming 400 cells in the x-direction

  # Create raster with equal horizontal and vertical resolution
  htrChina <- raster(
    xmn = xmin, xmx = xmax,
    ymn = ymin, ymx = ymax,
    res = cell_size,  # Specify resolution to be the same in both x and y
    crs = "+proj=longlat +datum=WGS84"
  )

  # Convert the raster to points to extract the coordinates
  htrChina_points <- rasterToPoints(htrChina, spatial = TRUE)

  # Set spatial coordinates for the input data frame
  coordinates(df) <- ~ Longitude_E + Latitude_N

  # Calculate distances between observation points and raster grid points
  dists <- spDists(
    x = coordinates(df),
    y = coordinates(htrChina_points),
    longlat = TRUE
  )

  # Inverse distance weighted interpolation
  idp <- 2
  inv.w <- (1 / (dists ^ idp))
  z <- (t(inv.w) %*% matrix(df$R_Sitting_ht)) / apply(inv.w, 2, sum)

  # Update the raster with predicted values
  htpred.China <- htrChina
  values(htpred.China) <- z

  # Export the data as a raster file in ASCII format
  writeRaster(htpred.China, filename = filename, format = "ascii", overwrite = TRUE)
}

# Apply the function to each group
# Interpolation for Early Female Population
perform_idw_interpolation(ht_china_male_early, "HT_IDW2_China_male_early.asc")

# Interpolation for Early Female Population
perform_idw_interpolation(ht_china_female_early, "HT_IDW2_China_female_early.asc")

# Interpolation for Late Male Population
perform_idw_interpolation(ht_china_male_late, "HT_IDW2_China_male_late.asc")

# Interpolation for Late Female Population
perform_idw_interpolation(ht_china_female_late, "HT_IDW2_China_female_late.asc")
```

```{r}
# Load the interpolated rasters
htpred_male_early <- raster("HT_IDW2_China_female_early.asc")
htpred_female_early <- raster("HT_IDW2_China_female_early.asc")
htpred_male_late <- raster("HT_IDW2_China_male_late.asc")
htpred_female_late <- raster("HT_IDW2_China_female_late.asc")

# Convert the rasters to data frames
htpred_male_early_df <- as.data.frame(htpred_male_early, xy = TRUE)
colnames(htpred_male_early_df) <- c("Longitude", "Latitude", "R_Sitting_ht")

htpred_female_early_df <- as.data.frame(htpred_female_early, xy = TRUE)
colnames(htpred_female_early_df) <- c("Longitude", "Latitude", "R_Sitting_ht")

htpred_male_late_df <- as.data.frame(htpred_male_late, xy = TRUE)
colnames(htpred_male_late_df) <- c("Longitude", "Latitude", "R_Sitting_ht")

htpred_female_late_df <- as.data.frame(htpred_female_late, xy = TRUE)
colnames(htpred_female_late_df) <- c("Longitude", "Latitude", "R_Sitting_ht")
```

```{r}
# Calculate global min and max for R_Sitting_ht across all datasets to unify the scale
global_min <- min(c(min(htpred_male_early_df $R_Sitting_ht, na.rm = TRUE),
                    min(htpred_female_early_df$R_Sitting_ht, na.rm = TRUE),
                    min(htpred_male_late_df$R_Sitting_ht, na.rm = TRUE),
                    min(htpred_female_late_df$R_Sitting_ht, na.rm = TRUE)))

global_max <- max(c(max(htpred_male_early_df$R_Sitting_ht, na.rm = TRUE),
                    max(htpred_female_early_df$R_Sitting_ht, na.rm = TRUE),
                    max(htpred_male_late_df$R_Sitting_ht, na.rm = TRUE),
                    max(htpred_female_late_df$R_Sitting_ht, na.rm = TRUE)))
```

```{r}
# Function to plot raster heatmap overlaid on the China base map with dataset points by Ethnic Group using shapes and colors
plot_heatmap_with_points_on_china_map <- function(raster_df, points_df, title, group_name) {
  # Determine the number of unique ethnic groups
  unique_ethnic_groups <- unique(points_df$Ethnic_Group)
  num_groups <- length(unique_ethnic_groups)

  # Define shape values for each ethnic group
  # Assign specific shapes to Han, Tibetan, and Sherpa groups
  shape_values <- rep(c(1:14), length.out = num_groups)
  names(shape_values) <- unique_ethnic_groups
  shape_values["Han"] <- 15     # Unify the Han group shape 
  shape_values["Tibetan"] <-19  # Unify the Tibetan group shape
  shape_values["Sherpa"] <- 17 # Assign specific shape for Sherpa group
  shape_values["Mongols"] <- 18 # Assign specific shape for Mongols group
  shape_values["Zhuang"] <- 20 # Assign specific shape for Zhuang group
  shape_values["Uyghurs"] <- 21 # Assign specific shape for Uyghurs group
  shape_values["Gelao"] <- 22 # Assign specific shape for Gelao group
  # Create the plot with unified layout settings
  p <- ggplot() +
    geom_sf(data = china_shape, fill = "white", color = "black") +  # Plot the base map of China
    geom_raster(data = raster_df, aes(x = Longitude, y = Latitude, fill = R_Sitting_ht), alpha = 0.7) +  # Plot the heatmap
    geom_point(data = points_df, aes(x = Longitude_E, y = Latitude_N, shape = Ethnic_Group, color = Ethnic_Group), size = 1) +  # Plot points using different shapes and colors
    scale_shape_manual(values = shape_values) +  # Assign custom shapes, including unified Han, Tibetan, and Sherpa group shapes
    scale_color_viridis_d(option = "plasma") +  # Assign colors using viridis color palette (option 'plasma' for variety)
    scale_fill_viridis_c(limits = c(global_min, global_max)) +  # Use viridis color scale for the heatmap with unified limits
    theme_minimal() +
    labs(title = title,
         x = "Longitude",
         y = "Latitude",
         fill = "Relative\nsitting height",
         shape = "Ethnic Group",
         color = "Ethnic Group",
         caption = group_name) +  # Add the group name under the plot
    theme(
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  # Center the title and make it bold
      axis.title = element_text(size = 10),  # Standardize axis title font size
      axis.text = element_text(size = 10),   # Standardize axis label size
      legend.position = "right",              # Move the legend (scale) to the left
      legend.title = element_text(size = 10),# Standardize legend title size
      legend.text = element_text(size = 8), # Standardize legend text size
      plot.margin = unit(c(1, 1, 1, 1), "cm") # Set consistent plot margins
    ) +
    coord_sf()  # Keep the coordinate reference consistent

  return(p)
}

# Plotting each dataset separately with the heatmap and the China base map

# Plot heatmap for Early Male Population with points
plot1 <- plot_heatmap_with_points_on_china_map(htpred_male_early_df, ht_china_male_early, "Heatmap with Points - Early Male Population", "Early Male Population")

# Plot heatmap for Early Female Population with points
plot2 <- plot_heatmap_with_points_on_china_map(htpred_female_early_df, ht_china_female_early, "Heatmap with Points - Early Female Population", "Early Female Population")

# Plot heatmap for Late Male Population with points
plot3 <- plot_heatmap_with_points_on_china_map(htpred_male_late_df, ht_china_male_late, "Heatmap with Points - Late Male Population", "Late Male Population")

# Plot heatmap for Late Female Population with points
plot4 <- plot_heatmap_with_points_on_china_map(htpred_female_late_df, ht_china_female_late, "Heatmap with Points - Late Female Population", "Late Female Population")
```
